Black-hole ringdown as a probe of higher-curvature gravity theories

### 2022-01-19 papers/paper_alt_theor_bounds/paper_alt_theor_bounds.tex
HS: reorganization of mod. grav. sec.

\documentclass[twocolumn,prd,aps,superscriptaddress,preprintnumbers,tightenlines,showpacs,nofootinbib,eqsecnum,amsfonts,amsmath,longbibliography]{revtex4-2}
\usepackage{epsfig}
\usepackage{graphics}
\usepackage{graphicx}
\usepackage{bm}
\usepackage[dvipsnames]{xcolor}
\usepackage{bm}
\usepackage{txfonts}
\usepackage{amssymb}
\usepackage{xspace}
\usepackage[normalem]{ulem} % To get strikethrough (\sout)
\usepackage[colorlinks]{hyperref}
\usepackage[caption=false]{subfig}
\usepackage{booktabs}
\usepackage{url}
\usepackage{float}
\usepackage[bottom]{footmisc}
\usepackage{lineno}
\usepackage{mathrsfs}
\usepackage{makecell}
\usepackage{microtype}

\definecolor{LinkColor}{rgb}{0.75, 0, 0}
\definecolor{CiteColor}{rgb}{0, 0.5, 0.5}
\definecolor{UrlColor}{rgb}{0, 0, 0.75}
\hypersetup{linkcolor=NavyBlue}
\hypersetup{citecolor=NavyBlue}
\hypersetup{urlcolor=NavyBlue}
\usepackage{perpage}
\MakePerPage{footnote}

@@

\newcommand{\pd}{\partial}
\newcommand{\dd}{{\rm d}}
\newcommand{\dV}{{\rm d}^{4}x \, \sqrt{-g} \,}
\newcommand{\lame}{\lambda_{\rm e}}
\newcommand{\lamo}{\lambda_{\rm o}}

@@

% HS: temporary title. Feel free to add suggestions
% \title{Probing higher-curvature gravity theories from binary black hole ringdown signals}
\title{The black-hole ringdown as a probe of higher-curvature gravity theories}

@@

% \hs{Here we review class of gravity theories we will consider.
% They all have higher-curvature terms, hence the `higher-curvature gravity theories'
% in the title.}

A broad class modified gravity theories are described by the action
%
\begin{equation}
S = \frac{1}{16 \pi G} \int \dd^4x \sqrt{-g}
\left[ R
- \tfrac{\beta}{2}(\nabla \varphi)^2
\right]
+
\sum_{n \geqslant 2} \frac{\ell^{2n - 2}}{16 \pi G} S^{(2n)},
\end{equation}
%
where $\varphi$ is a dynamical scalar field, $\ell$ some length scale, assumed
to be small compares to the length scale associated with a black hole, i.e.,
$GM \gg \ell$, and $S^{(2n)}$ are corrections to the Einstein-Klein-Gordon
action that introduce involving higher-order curvature tensors (with $2n$
metric derivatives) and scalar field.

More specifically, we work up to dimension-eight operators (i.e., $n=4$),
%
\begin{subequations}
\begin{align}
S^{(4)} &= \alpha_{\rm GB} R^{2}_{\rm GB} + \alpha_{\rm CS} {}^{*}RR,
\label{eq:s_4}
\\
S^{(6)} &= \lame R_{\mu\nu}{}^{\rho\sigma} R_{\rho\sigma}{}^{\gamma\delta} R_{\gamma \delta}{}^{\mu\nu}
+ \lamo \ell^4 R_{\mu\nu}{}^{\rho\sigma} R_{\rho\sigma}{}^{\gamma\delta} \tilde{R}_{\gamma \delta}{}^{\mu\nu},
\label{eq:s_6}
\\
S^{(8)} &= \epsilon_{1} \mathcal{C}^2  + \epsilon_{2} \tilde{\mathcal{C}}^{2}
+ \epsilon_{3} \mathcal{C} \tilde{\mathcal{C}},
\label{eq:s_8}
\end{align}
\label{eq:action_mods}
\end{subequations}
%
where we defined the Gauss-Bonnet and Pontryagin densities, respectively,
%
\begin{subequations}
\begin{align}
    R^{2}_{\rm GB} &= R_{\mu\nu\rho\sigma}R^{\mu\nu\rho\sigma} - 4 R_{\mu\nu} R^{\mu\nu} + R^2,
    \\
    {}^{*}RR &= \tfrac{1}{2} R_{\nu\mu\rho\sigma} {}^{*}R^{\mu\nu\rho\sigma},
\end{align}
\end{subequations}
%
and where $R_{\mu\nu\rho\sigma}$ is the Riemann tensor
and $\tilde{R}_{\mu\nu\rho\sigma}$ its dual, defined as
%
\begin{equation}
\tilde{R}_{\mu\nu\alpha\beta} = \tfrac{1}{2} \epsilon_{\mu\nu\rho\sigma} R^{\rho\sigma}{}_{\alpha\beta},
\end{equation}
%
with $\epsilon_{\mu\nu\rho\sigma}$ being the Levi-Civita tensor.

This action encapsulates various modified gravity theories, which often are studied on their own,
namely,
%
\begin{itemize}
    \item scalar Gauss-Bonnet gravity (sGB): corresponds to $\alpha_{\rm GB}$
          nonzero and all other dimensionless coupling constants equal to zero.
    \item dynamical Chern-Simons (dCS): this corresponds to $\alpha_{\rm CS}$
          nonzero and all other dimensionless coupling constants equal to zero.
    \item cubic effective field theory of GR (cEFTofGR): this corresponds to $\lame$, $\lamo$ nonzero
          and all other dimensionless coupling constants equal to zero.
    \item quartic effective field theory of GR (qEFTofGR): this corresponds to $\varepsilon_{i}$
          ($i = 1,2,3$) nonzero and all other dimensionless coupling constants equal to zero.
\end{itemize}

In each of these theories, slowly-rotating black hole and their quasinormal mode spectra
to leading order in spin have been calculated.
%
We can use this results to establish a mapping between theory-specific calculations and
the theory-agnostic {\sc Parspec} parametrization.
%
We discuss how we do this in the section.

In Eqs.~\eqref{eq:action_mods}, $\alpha_{\rm GB}$, $\alpha_{\rm CS}$, $\lame$,
$\lamo$ and $\varepsilon_{i}$ are dimensionless constants.

>>>>

\documentclass[twocolumn,prd,aps,superscriptaddress,preprintnumbers,tightenlines,showpacs,nofootinbib,eqsecnum,amsfonts,amsmath,longbibliography]{revtex4-2}
\usepackage{epsfig}
\usepackage{graphics}
\usepackage{graphicx}
\usepackage[dvipsnames]{xcolor}
\usepackage{bm}
% \usepackage{txfonts}
\usepackage{amssymb}
\usepackage{xspace}
\usepackage[normalem]{ulem} % To get strikethrough (\sout)
\usepackage[colorlinks]{hyperref}
\usepackage[caption=false]{subfig}
\usepackage{booktabs}
\usepackage{url}
\usepackage{float}
\usepackage[bottom]{footmisc}
\usepackage{lineno}
\usepackage{mathrsfs}
\usepackage{makecell}
\usepackage{microtype}

\usepackage{tikz}

\definecolor{LinkColor}{rgb}{0.75, 0, 0}
\definecolor{CiteColor}{rgb}{0, 0.5, 0.5}
\definecolor{UrlColor}{rgb}{0, 0, 0.75}
\definecolor{rred}{RGB}{218,41,28}
\hypersetup{linkcolor=NavyBlue}
\hypersetup{citecolor=NavyBlue}
\hypersetup{urlcolor=NavyBlue}

\usepackage{perpage}
\MakePerPage{footnote}

@@

% Modified gravity related
\newcommand{\pd}{\partial}
\newcommand{\dd}{{\rm d}}
\newcommand{\dV}{{\rm d}^{4}x \, \sqrt{-g} \,}

\newcommand{\lame}{\lambda_{\rm e}}
\newcommand{\lamo}{\lambda_{\rm o}}
\newcommand{\acs}{\ell_{\rm CS}}
\newcommand{\agb}{\ell_{\rm GB}}

@@

% HS: temporary title. Feel free to add suggestions
% \title{Probing higher-curvature gravity theories from binary black hole ringdown signals}
\title{Black-hole ringdown as a probe of higher-curvature gravity theories}

@@

We will consider several modified gravity theories as applications of the \pSEOB{}
waveform model presented in Sec.~\ref{sec:review_pSEOB}.
%
In the following, we briefly review which theories we consider, what are the
currents observational constraints in each of them and what we known about the
QNMs of black holes in each of them.

\subsubsection{scalar-Gauss-Bonnet gravity}
%
This theory is given by the following action,
%
\begin{equation} \label{eq:action_sgb}
    S_{\rm GB} = \frac{1}{16 \pi}
    \int \dV
    \left[
    R - \tfrac{1}{2}(\nabla \varphi)^2
    + \tfrac{1}{4} \ell^{2}_{\rm GB} f(\varphi) \mathscr{G}
    \right],
\end{equation}
%
where $R$ is the Ricci scalar associated with the metric $g_{\alpha\beta}$ and
$\varphi$ is a dynamical scalar field which couples to the Gauss-Bonnet
invariant $\mathscr{G}$,
%
\begin{equation} \label{eq:def_gb}
    \mathscr{G} =
    R^{\mu\nu\rho\sigma}R_{\mu\nu\rho\sigma}
    - 4 R^{\mu\nu}R_{\mu\nu}
    + R^2,
\end{equation}
%
with strength set by the dimensionful coupling constant $\agb$ with dimensions
of length.

Different subclasses of this theory are determined by the function $f(\varphi)$
and they can be classified into two classes based on the properties of their black hole solutions.
%
In the first class, the first derivative of the coupling
function $f'(\varphi) = \dd f  / \dd \varphi$ is always nonzero and black holes
are known to always support (secondary) scalar hair.
%
Examples include the shift-symmetric $f \propto \varphi$ and dilatonic
$f \propto \exp(\varphi)$ couplings.
%
In the second class, $f'(\varphi) = 0$ can vanish for some constant $\varphi_0$.
%
In this case, one can show that theory admits the same stationary,
asymptotically flat black hole solution of GR and scalarized black
holes~\cite{Doneva:2017bvd,Silva:2017uqg,Dima:2020yac,Herdeiro:2020wei,Berti:2020kgk}.
%
Examples include the quadratic $f \propto \varphi^2$~\cite{Silva:2017uqg}
and Gaussian $f \propto \exp(-\varphi^2)$~\cite{Doneva:2017bvd} couplings.
%
Here will consider only the dilatonic theory.

Black holes in both classes support monopolar scalar hair and thus, when in binaries,
can source scalar dipole radiation and are therefore prone to be constrained with
GWs observations of compact binaries~\cite{Nair:2019iur,Perkins:2021mhb}.
%
\dots

The calculation of the QNMs in this theory is more complicated than in GR already for
nonrotating black holes. The reason is due to the coupling between scalar field and
the Gauss-Bonnet invariant, which manifests into a coupling between scalar perturbation
and gravitational perturbations of polar parity \dots

\begin{figure}
\begin{tikzpicture}[baseline,>=stealth]
    % \draw[help lines, gray] (-4, -2) grid (4, 2);
    \draw[->] (-2.5, 0) -- (-1.75, +0.5);
    \draw[->] (-2.5, 0) -- (-1.75, -0.5);

    \node at (-3.25, 0) {$\{h_{\alpha\beta}, \, \varphi\}$};

    \node at (-1, 1.25) {${\cal O}(\chi^{0})$};
    \node at (-1, +0.5) {$\{ \textcolor{rred}{h^{\rm p}_{\alpha\beta}}, \, \varphi\}$};
    \node at (-1, -0.5) {$h^{\rm a}_{\alpha\beta}$};

    \node at (+1, 1.25) {${\cal O}(\chi^{1})$};
    \node at (+1, +0.5) {\dots};
    \node at (+1, -0.5) {\dots};
\end{tikzpicture}
\caption{Quasinormal modes of black holes in scalar-Gauss-Bonnet gravity.}
\end{figure}

\subsubsection{dynamical Chern-Simons}

This theory is given by the following action,
%
\begin{equation} \label{eq:action_dcs}
    S_{\rm CS} = \frac{1}{16 \pi}
    \int \dV
    \left[
    R - \tfrac{1}{2}(\nabla \vartheta)^2
    + \tfrac{1}{4} \ell^{2}_{\rm CS} \, \vartheta \, {}^{*}RR
    \right],
\end{equation}
%
where $\vartheta$ is a pseudo-scalar field, which couples
to the Pontryagin density,
%
\begin{equation}
    {}^{*}RR = {}^{*}R^{\mu}{}_{\nu}{}^{\rho\sigma} R^{\nu}_{\mu\rho\sigma},
\end{equation}
%
where ${}^{*}R^{\mu}{}_{\nu}{}^{\rho\sigma}$ is the dual of the Riemann tensor,
defined as
%
${}^{*}R^{\mu}{}_{\nu}{}^{\rho\sigma} =
\tfrac{1}{2} \epsilon^{\mu}{}_{\nu\gamma\delta}
R^{\gamma\delta\rho\sigma}$,
%
and where $\epsilon^{\mu}{}_{\nu\gamma\delta}$ is the Levi-Civita tensor.
%
The coupling between scalar field and the Pontryagin density is set by the
$\ell_{\rm CS}$ with dimensions of length.

The theory admits the Schwarzschild black hole solution of GR, but predicts
that spinning black holes have (secondary) scalar hair, which, unlike in
scalar-Gauss-Bonnet gravity, is dipolar~\cite{Yunes:2009hc,Konno:2009kg}.
%
Consequently, the leading-scalar radiation channel is quadrupolar making this
theory presently unconstrained by from the GWs from the inspiral of black hole
binaries alone~\cite{Nair:2019iur,Perkins:2021mhb}.
%
However, the theory has been constrained by Ref~\cite{Silva:2020acr} through a
combination information of x-ray observations of the isolated neutron star PSR
J0030+0451~\cite{Lommen:2000yt,NANOGrav:2017wvv} by the
NICER~\cite{Riley:2019yda,Miller:2019cac} and gravitational-wave observation of
the binary neutron star event GW170817~\cite{TheLIGOScientific:2017qsa}.

Similarly to the case of scalar-Gauss-Bonnet gravity, the perturbations of nonrotating
black holes in dynamical Chern-Simons gravity also have a coupling between scalar and
metric perturbations. However, the coupling is between the scalar and gravitational perturbations
of axial parity~\cite{Yunes:2007ss,Cardoso:2009pk,Molina:2010fb,Wagle:2021tam}.

\begin{figure}
\begin{tikzpicture}[baseline,>=stealth]
    % \draw[help lines, gray] (-4, -2) grid (4, 2);
    \draw[->] (-2.5, 0) -- (-1.75, +0.5);
    \draw[->] (-2.5, 0) -- (-1.75, -0.5);

    \node at (-3.25, 0) {$\{h_{\alpha\beta}, \, \vartheta\}$};

    \node at (-1, 1.25) {${\cal O}(\chi^{0})$};
    \node at (-1, +0.5) {$h^{\rm p}_{\alpha\beta}$};
    \node at (-1, -0.5) {$\{ \textcolor{rred}{h^{\rm a}_{\alpha\beta}}, \, \vartheta\}$};

    \node at (+1, 1.25) {${\cal O}(\chi^{1})$};
    \node at (+1, +0.5) {\dots};
    \node at (+1, -0.5) {\dots};
\end{tikzpicture}
\caption{Quasinormal modes of black holes in dynamical Chern-Simons gravity.}
\end{figure}

\subsubsection{Effective-field-theory of GR}

This theory is given by the following action,
%
\begin{equation} \label{eq:action_eft}
    S_{\rm EFT} = \frac{1}{16 \pi}
    \int \dd^4x \sqrt{-g}
    \left[ R
    +
    \footnotesize{\sum}_{n \geqslant 2} \ell^{2n - 2} L^{(2n)}
    \right],
\end{equation}
%
where $\ell$ is some length scale, assumed to be small compares to the length
scale associated with a black hole, i.e., $\ell / M \ll 1$, and
$L^{(2n)}$ are corrections to the Einstein-Hilbert term action that
introduce involving higher-order curvature tensors (with $2n$ metric
derivatives).

More specifically, we follow Refs.~\cite{Cano:2020cao,Cano:2021myl} and work up
to dimension-eight operators (i.e., $n=4$),
%
\begin{subequations}
\label{eq:action_mods}
\begin{align}
    L^{(6)} &= \lame R_{\mu\nu}{}^{\rho\sigma} R_{\rho\sigma}{}^{\gamma\delta} R_{\gamma \delta}{}^{\mu\nu}
    \nonumber \\
            &\quad + \lamo R_{\mu\nu}{}^{\rho\sigma} R_{\rho\sigma}{}^{\gamma\delta} \tilde{R}_{\gamma \delta}{}^{\mu\nu},
    \label{eq:s_6}
    \\
    L^{(8)} &= \varepsilon_{1} \mathcal{C}^2
    + \varepsilon_{2} \tilde{\mathcal{C}}^{2}
    + \varepsilon_{3} \mathcal{C} \tilde{\mathcal{C}},
\label{eq:s_8}
\end{align}
\end{subequations}
%
where $\lambda_{\rm o, e}$ and $\varepsilon_{i}$ ($i=1,2,3$) are dimensionless parameters.

Black holes in these theories were studied in \dots

% where we defined the Gauss-Bonnet and Pontryagin densities, respectively,
% %
% \begin{subequations}
% \begin{align}
%     R^{2}_{\rm GB} &= R_{\mu\nu\rho\sigma}R^{\mu\nu\rho\sigma} - 4 R_{\mu\nu} R^{\mu\nu} + R^2,
%     \\
%     {}^{*}RR &= \tfrac{1}{2} R_{\nu\mu\rho\sigma} {}^{*}R^{\mu\nu\rho\sigma},
% \end{align}
% \end{subequations}
% %
% and where $R_{\mu\nu\rho\sigma}$ is the Riemann tensor
% and $\tilde{R}_{\mu\nu\rho\sigma}$ its dual, defined as
% %
% \begin{equation}
% \tilde{R}_{\mu\nu\alpha\beta} = \tfrac{1}{2} \epsilon_{\mu\nu\rho\sigma} R^{\rho\sigma}{}_{\alpha\beta},
% \end{equation}
% %
% with $\epsilon_{\mu\nu\rho\sigma}$ being the Levi-Civita tensor.

% This action encapsulates various modified gravity theories, which often are studied on their own,
% namely,
% %
% \begin{itemize}
%     \item scalar Gauss-Bonnet gravity (sGB): corresponds to $\alpha_{\rm GB}$
%           nonzero and all other dimensionless coupling constants equal to zero.
%     \item dynamical Chern-Simons (dCS): this corresponds to $\alpha_{\rm CS}$
%           nonzero and all other dimensionless coupling constants equal to zero.
%     \item cubic effective field theory of GR (cEFTofGR): this corresponds to $\lame$, $\lamo$ nonzero
%           and all other dimensionless coupling constants equal to zero.
%     \item quartic effective field theory of GR (qEFTofGR): this corresponds to $\varepsilon_{i}$
%           ($i = 1,2,3$) nonzero and all other dimensionless coupling constants equal to zero.
% \end{itemize}
%
% In each of these theories, slowly-rotating black hole and their quasinormal mode spectra
% to leading order in spin have been calculated.
% %
% We can use this results to establish a mapping between theory-specific calculations and
% the theory-agnostic {\sc Parspec} parametrization.
% %
% We discuss how we do this in the section.
%
% In Eqs.~\eqref{eq:action_mods}, $\alpha_{\rm GB}$, $\alpha_{\rm CS}$, $\lame$,
% $\lamo$ and $\varepsilon_{i}$ are dimensionless constants.
